\input{notes.tex}

\iftoggle{dualscreen}{\setbeameroption{show notes on second screen=right}}{}

\begin{document}
\section{Lecture 3}
\subsection{Review}



\slide[Review]{
\itmz{
\item Linear DEs: $y\p +p(t) y = g(t); \qquad y(t_0)=y_0$
\itmz{
\item \alert{Homogeneous}: \student{$g(t)=0$}
\item \alert{Inhomogeneous}:\student{ $g(t)$ is not zero everywhere.}
\item \alert{Constant coefficient} $p(t)=a$, with constant $a$.
\student{\itmz{
\item General solution of form: $y_g=y_h+y_p$
\item $y_h = C e^{-at}$
\item $y_p$ – Method of undetermined coefficients\\\hspace{4em}-Skipped, will revisit
}
}
}
}
\vfill
What to do if $p(t)$ is not constant, or if undetermined coefficients doesn't work for a specific $g(t)$?
\student{\subitem{Use the method of integrating factors.}}
}
\subsection{Preamble}
\slide[Want to solve $y\p+p(t)y=g(t)$]{
Instead, lets do something easier. Solve \[\dd{}{t}\left(q \cdot y \right) = h(t)\] with some known $q(t)$ and $\int h(t) dt = H(t)+C$.

\student{\algn{\dd{}{t}\left(q(t) \cdot y(t) \right) &= h(t) & \rightarrow && \text{d} \left(q \cdot y \right)&= h(t) \text{d}t\\
q(t)y(t)& = H(t)+C &  \leftarrow && \int \text{d} \left(q(t)y(t)\right) &= \int h(t) \text{d}t}
\vfill
\[\Rightarrow y(t) = \frac{H(t)+C}{q(t)}\]
}

}



\slide[Idea]{
It would be very nice if we could multiply \[ y\p + p(t) y = g(t)\] by some magical function $\mu(t)$ such that \[\mu(t)\cdot \left[y\p + p(t) y\right] =\dd{}{t}\left[q(t)\cdot y(t)\right]  \]
\vfill
\student{Unique choice of $\mu(t)$:\[\mu(t)=e^{\intop p(t)dt}\]
\vfill See full notes on Canvas for details. }
\vfill
}



\subsection{Method of integrating factors}
\slide[Method of integrating factors: $y\p+p(t)y=g(t) $ ]{
\enum{
\item Multiply by the integrating factor $\mu(t)=e^{\intop p(t)\text{d}t  }$ \student{\algn{\mu(t)y\p +\underbrace{p(t)\mu(t)}_{\mu\p(t)}y &= g(t)\mu(t) \\
\dd{}{t}(\mu \cdot y) &= g(t)\mu(t)\\
\text{d}(\mu \cdot y) &= g(t)\mu(t)\text{d}t
}}
\item Integrate: \student{\[\mu \cdot y = \intop g(t)\mu(t)\text{d}t + C \]}

\item Isolate $y(t)$: \student{\[ y(t) = \frac{\intop g(t)\mu(t)\text{d}t + C}{\mu(t)} \]}

}\vfill
\student{All boils down to evaluating two integrals, no geussing.}
}
\subsection{Method of integrating factors: $y\p+p(t)y=g(t) $}
\slide[Existence and Uniqueness \hspace{4em} $\rarray{ y(t) = \frac{\intop g(t)\mu(t)\text{d}t + C}{\mu(t)} \\ \mu(t)=e^{\intop p(t)\text{d}t }}$]{
\itmz{
\item What properties does $p(t)$ need to ensure $\mu(t)$ is differentiable?
\student{\subitem{Continuous and integrable.}}
\vfill
\item What properties does $g(t)$ need to ensure $y(t)$ is differentiable?
\student{\subitem{Continuous and $\mu(t) \cdot g(t)$ is integrable.}
\vfill
These conditions are pretty lax:\subitem{ almost all linear 1st Order DEs can be solved.}
\vfill
Evaluating the integrals can be tough.
}
}
}

\slide{\ex{Find the general solution of $y\p-3y=e^t$.} \hfill $\boxed {\rarray{ y(t) = \frac{\intop g(t)\mu(t)\text{d}t + C}{\mu(t)} \\ \mu(t)=e^{\intop p(t)\text{d}t }}}$

\student{\algn{\mu(t)&=e^{\intop(-3)dt}=e^{-3t}\\
e^{-3t}y\p-3e^{-3t}y&=e^te^{-3t}\\
\dd{}{t}\left( e^{-3t}y \right ) &=e^{-2t}\\
\text{d}\left( e^{-3t}y \right ) &=e^{-2t}\text{d}t\\
e^{-3t}y(t)  &= \intop e^{-2t}\text{d}t+C\\
y(t) &= \frac{ -\frac12 e^{-2t}+C}{e^{-3t}}\\
&=\underbrace{-\frac12e^t}_{y_p}+\underbrace{Ce^{3t}}_{y_h}
}
}

}

\slide{\ex{Find the general solution of $y\p-3t^2y=3t^2$.}  \hfill $\boxed {\rarray{ y(t) = \frac{\intop g(t)\mu(t)\text{d}t + C}{\mu(t)} \\ \mu(t)=e^{\intop p(t)\text{d}t }}}$
\vfill\student{\algn{\mu(t)&=e^{-\intop3t^2dt}=e^{-t^3} &                                                                                    & \carray{\text{let }u=t^3\\du=3t^2}\\
e^{-t^3}y(t)  &= \intop3t^2e^{-t^3}\text{d}t&                                                          \intop 3t^2e^{-t^3}\text{d}t & = \intop e^{-u} du\\
                       &                                                                                                                                                             &   & = -e^{-u}+C\\
e^{-t^3}y(t) &= -e^{-t^3}+C\\
y(t)&=\underbrace{-1}_{y_p}+\underbrace{Ce^{t^3}}_{y_h}
}
}
}

\slide[Solution Structure for $y\p+ay=g(t)$]{
Do we always have a general solution of the form $y_g=y_p+y_h$?
\vfill
\cstudent{Yes, always.\vfill For all \uline{linear} DEs.}
}

\slide[Solution Structure  for $y\p+p(t)y=g(t)$]{
Do we always have a general solution of the form $y_g=y_p+y_h$?

Suppose $y_p$ and $y_h$ solve the following DEs \algnlbl{y_p\p+p(t)y_p&=g(t)\\y_h\p+p(t)y_h&=0}

What DE does $y_g$ solve?
\student{\algn{y_g\p+p(t)y_g &= y_p\p+y_g\p+p(t)(y_p+y_h) \\ &  =\left(y_p\p+p(t)y_p\right) +\left(y_h\p+p(t)y_h\right) \\&=g(t) + 0 = g(t)} \vfill $y_g$ also solves Eq. (1)! Give me some $y_p$ that solves (1), I can always add $y_h$ and still have a solution to (1).}
}


\slide[Solution Structure  for $y\p+p(t)y=g(t)$]{\vspace{-1em}
Do we always have a general solution of the form $y_g=y_p+y_h$?

Suppose $y_p$ solves the following DEs \algn{y_p\p+p(t)y_p&=g(t) && \hfill (1) }

Is $y_p$ necessarily a particular solution?\student{Proof by contradiction\\Assume $y_p$ has some arbitrary constant.\vfill

\twomini[.4]{.5}{.5}{
Additive Constant: $y_p\rightarrow y_p+C$ \algn{y_p\p+a(y_p+C)&=y_p\p+ay_p+aC\\&= g(t)+aC \\&\neq g(t)}
}{
Multiplicative Constant: $y_p\rightarrow C y_p$ \algn{Cy_p\p+aCy_p&=C\left(y_p\p+ay_p\right)\\&= Cg(t) \\&\neq g(t)}
}
$y_p$ cannot have any arbitrary constants!
}
}

\slide[Summary]{
\itmz{
\item Looked at general linear 1st order DEs: $y\p +p(t)y=g(t)$
\itmz{
\item Tried to turn the LHS into an exact derivative: $$y\p +p(t)y \quad \rightarrow \quad \text{d}(q\cdot y ) = q\p y +q y\p$$
\subitem{ Multiplied by an integrating factor $\mu(t)$
\item After some algebra find that $\mu=q$
\item General solution: $ y(t) = \frac{\intop g(t)\mu(t)\text{d}t + C}{\mu(t)}$
\item Integrating factor: $\mu(t)=e^{\intop p(t)\text{d}t }$}

}
\item Found that we could solve all linear 1st order DEs by this method, provided e.g. that $p(t)$and $g(t)$ are continuous.
\item General solution to a linear DE: $y_g = y_p +y_h$
}
}
\end{document}